---
title:  "Pearl’s backdoor criterion"
author: "Beate Sick"
date:   "`r Sys.Date()`"
editor: source
lang: de
language: 
  title-block-author-single: "Autor:"
  title-block-published: "Version vom:"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    theme: cosmo
    fig-width: 8
    fig-height: 6
    fig-align: "center"
    fig-caption: true
    df-print: kable
    highlight: "monochrome"
---

```{css, echo = FALSE}
.justify {
  text-align: justify
}
```

## Exercise 1  {.justify}

The shown causal DAG corresponds to a known data generating process (linear SCM) which has been used
to simulate observational data for the involved variable X1, X2, X3, X4, X5.
```{r, message=FALSE, warning=FALSE}
library(dagitty)
library(tidyverse)
library(ggdag)
g <- dagitty('dag {
             X3 -> {X2 X1 X5}
             X1 -> X4 
             X2 -> X5
             X4 -> X5
             
            
}')
set.seed(8173)
ggdag(g, 
      layout = "nicely", 
      text_col = "black", 
      #label_col = "blue", 
      node = F, 
      stylized = T) +
  geom_dag_point(color = "blue", alpha = 0.3) +
  theme_dag_blank()
```

### Teilaufgabe a
a) Has X1 (“treatment”) potentially a causal effect on   
(i) X5 (“outcome 1”)?  [yes, there is a direct causality between X1 and X5]{style="color:#00008B; font-weight:bold"}
(ii) X2 (“outcome 2”)?  [No, there is no direct causality between X1 and X2]{style="color:#00008B; font-weight:bold"}

```{r}
adjustmentSets(g, exposure = "X1", outcome = "X2")
```
```{r}
adjustmentSets(g, exposure = "X1", outcome = "X5")
```

### Teilaufgabe b
se Pearl’s backdoor (BD) criterion to assess in which of the following regression models the coefficient
of X1 gives an unbiased estimate of the total causal effect (=intervention effect) of X1 on X5.  
$(i) X5 ∼ X1$     
$(ii) X5 ∼ X1 + X2$  
$(iii) X5 ∼ X1 + X3$  
$(iv) X5 ∼ X1 + X2 + X4$  
$(v) X5 ∼ X1 + X2 + X3$  

```{r}
adjustmentSets(g, exposure = "X1", outcome = "X5", effect = "total", type = "all")

```
[Thus the correct model is $(iii) X5 ∼ X1 + X3$, $(i)$ & $(ii)$ & $(iv)$ would give a biased estimate of the total causal effect of X1 on X5. $(v) X5 ∼ X1 + X2 + X3$ would give an unbiased estimate of the total causal effect of X1 on X5, but it is not the minimal adjustment set.]{style="color:#00008B; font-weight:bold"}


### Teilaufgabe c
Use the provided simulated data (sim.csv) to fit the 5 proposed regressions with linear regression and
determine a 95% CI for the estimated coefficient of X1.  
$(i) X5 ∼ X1$  
$(ii) X5 ∼ X1 + X2$  
$(iii) X5 ∼ X1 + X3$  
$(iv) X5 ∼ X1 + X2 + X4$  
$(v) X5 ∼ X1 + X2 + X3$  

```{r}
library(readr)
sim <- read_delim("sim.csv", delim = ";", 
    escape_double = FALSE, trim_ws = TRUE)
sim <- as.data.frame(sim)
```

$(i) X5 ∼ X1$ 
```{r}
fit.i <- lm(x5 ~ x1, data = sim)
(confint(fit.i))


```
$(ii) X5 ∼ X1 + X2$ 
```{r}
fit.ii <- lm(x5 ~ x1 + x2, data = sim)
(confint(fit.ii))
```

$(iii) X5 ∼ X1 + X3$  
```{r}
fit.iii <- lm(x5 ~ x1 + x3, data = sim)
(confint(fit.iii))
```

$(iv) X5 ∼ X1 + X2 + X4$  
```{r}
fit.iv <- lm(x5 ~ x1 + x2 + x4, data = sim)
(confint(fit.iv))
```
$(v) X5 ∼ X1 + X2 + X3$  
```{r}
fit.v <- lm(x5 ~ x1 + x2 + x3, data = sim)
(confint(fit.v))
```

### Teilaufgabe d
Look at the given SCM defining the data generating Process. Draw the corresponding DAG and
indicate the linear coefficients at the edges.  
$\epsilon_i ∼ N (0, 1)$  
$X3 ← \epsilon_3$  
$X1 ← 5 × X3 + \epsilon_1$  
$X2 ← 9 × X3 + \epsilon_2$  
$X4 ← 7 × X1 + \epsilon_4$  
$X5 ← 5 × X2 + 8 × X3 − 9 × X4 + \epsilon_5$  

```{r, message=FALSE, warning=FALSE}
g <- dagitty('dag {
             X1 [pos="0,1"]
             X2 [pos="2,2"]
             X3 [pos="0,2"]
             X4 [pos="1,0"]
             X5 [pos="2,1"]
            
             X1 -> X4 [beta = 7] 
             X2 -> X5 [beta = 5]
             X3 -> X1 [beta = 5]
             X3 -> X2 [beta = 9]
             X3 -> X5 [beta = 8]
             X4 -> X5 [beta = -9]
             
            
}')
set.seed(8173)
plot(g, show.coefficients = TRUE)

ggdag(g, 
      layout = "nicely", 
      text_col = "black", 
      label_col = "blue", 
      node = F, 
      stylized = T,
      use_labels = NULL,
      text = T) +
  geom_dag_point(color = "blue", alpha = 0.3) +
  theme_dag_blank()
```
