---
title:  "Dagitty"
author: "Beate Sick"
date:   "`r Sys.Date()`"
editor: source
lang: de
language: 
  title-block-author-single: "Autor:"
  title-block-published: "Version vom:"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    theme: cosmo
    fig-width: 8
    fig-height: 6
    fig-align: "center"
    fig-caption: true
    df-print: kable
    highlight: "monochrome"
---

```{css, echo = FALSE}
.justify {
  text-align: justify
}
```

## Exercise 1  {.justify}

In this tasks, the Structural Causal Models (SCM) for the data generating process are given. The goal of
the tasks is to draw the corresponding causal graph and perform a simulation study.
Excercise 1
Consider the following SCM:
U1 ∼ N (0, 1)
U2 ∼ N (0, 1)
U3 ∼ N (0, 1)
X3 = U3
X1 = 0.7 · X3 + U1
X2 = 1 · X1 + 0.5 · X3 + U2

### Teilaufgabe a

Use Daggity to plot the corresponding DAG.
```{r}
library(dagitty)
g <- dagitty('dag {
             X3 -> {X2 X1}
             X1 -> X2
            
}')

plot(g)
```


### Teilaufgabe b
Simulate n = 1000 observations according to this SCM (set.seed(1)) and save the observed variables in a data.frame called “dat” and visualize the data. Generate a histogram of the observed variable x2 and determine its mean and median. Compute a 95% confidence interval for the expected value of X2.

```{r}
set.seed(1)
n <-  1000
u1 <- rnorm(n, mean = 0, sd = 1)
u2 <- rnorm(n, mean = 0, sd = 1)
u3 <- rnorm(n, mean = 0, sd = 1)

X3 <- u3
X1 <- 0.7 * X3 + u1
X2  <- X1 + 0.5 * X3 + u2

dat <- data.frame(u1 = u1, u2 = u2, u3 = u3, X1 = X1, X2 = X2, X3 = X3)
head(dat, 3)
```
```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(GGally)
ggpairs(dat)

```

```{r}
hist(dat$X2)
```


```{r}
mean(dat$X2)
median(dat$X2)
t <- t.test(dat$X2,conf.level =  0.95)
t
```


```{r}
t$conf.int
```

### Teilaufgabe c
Assume that X3 is increased by one unit. Write down the updated SCM (replace in all equations X3
by (X3+1)) and simulate again n=1000 observations from the update SCM and save the simulated
data into a data.frame called dat1. Determine the mean of x2 and the 95% CI for E(X2) and compare
it with the corresponding results that you have received with the original SCM. Compute difference of
the two point estimates for E(X2).


```{r}
set.seed(1)
n <-  1000
u1 <- rnorm(n, mean = 0, sd = 1)
u2 <- rnorm(n, mean = 0, sd = 1)
u3 <- rnorm(n, mean = 0, sd = 1)

X3 <- u3
X1 <- 0.7 * (X3 + 1) + u1
X2  <- X1 + 0.5 * (X3 + 1) + u2

dat1 <- data.frame(u1 = u1, u2 = u2, u3 = u3, X1 = X1, X2 = X2, X3 = X3)
head(dat1, 3)
```
```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(GGally)
ggpairs(dat1)

```

```{r}
hist(dat1$X2)
```


```{r}
mean(dat1$X2)
median(dat1$X2)
t <- t.test(dat1$X2,conf.level =  0.95)
t
```


```{r}
t$conf.int
```

### Teilaufgabe d
How large is the direct causal effect of X3 on X2 according to the SCM? 
How large is the total causal effect of X3 on X2 according to the SCM? 
Tipp: Recall the path rule for linear SCMs.

X3 <- u3
X1 <- 0.7 * (X3 + 1) + u1
X2  <- X1 + 0.5 * (X3 + 1) + u2

direct causal effect of X3 on X2: 0.5 x (X3 + +) = 0.5 x 1 = 0.5
total causal effect X3 on X2: X1 + 0.5 x (X3 + 1) = 0.7 x (X3 + 1) + 0.5 x (X3 + 1) = 0.7 + 0.5 = 1.2

### Teilaufgabe e
Use the simulated data from the original SCM and fit two different linear regression models, both with
X2 as target: 1) X2~X3 , 2) X2~X3+X1. Determine the CIs for the estimated coefficients and discuss
the results.
```{r}
fit1 <- lm(X2 ~ X3, data = dat)
summary(fit1)
confint(fit1)
```
[In this unajusted model we get the coefficient $\beta_3$ that corresponds to the total causal effect of X3 on X2. The confidence interval for $\beta_3$ is $(1.18, 1.36)$.]{style="color:#00008B; font-weight:bold"}

```{r}
fit2 <- lm(X2 ~ X3 + X1, data = dat)
summary(fit2)
confint(fit2)
```
[In this model, where we adjusted for X1,  we get the coefficient $\beta_3$ that corresponds to the direct causal effect of X3 on X2. The confidence interval for $\beta_3$ is $(0.44, 0.59)$.]{style="color:#00008B; font-weight:bold"}

### Teilaufgabe f
Use daggity to ask for the correct adjustment set to get the total or direct causal effect of X3 on X2.
```{r}
adjustmentSets(g, exposure = "X3", outcome = "X2", effect = "total")
```
[dagitty returns the empty set, which means that there is no adjustment set that can be used to estimate the total causal effect of X3 on X2.]{style="color:#00008B; font-weight:bold"}

```{r}
adjustmentSets(g, exposure = "X3", outcome = "X2", effect = "direct")
```
[dagitty returns the set {X1}, which means that X1 is the correct adjustment set to estimate the direct causal effects of X3 on X2.]{style="color:#00008B; font-weight:bold"}


# Exercise 2
## Teilaufgabe a
What is the total cause of X on Y

1

## Teilaufgabe b
Simulate n = 1000 data points. Then fit (1) a simple regression with X as the only predictor and (2)
a multiple linear regression with both X and Z as predictors. In which of the fitted regression models
is the true causal effect within the confidence interval of the effect of X on Y ? Set your seed to 123.
```{r}
set.seed(123)
n <-  1000
Ex <- rnorm(n, mean = 0, sd = 1)
Ey <- rnorm(n, mean = 0, sd = 1)
Ez <- rnorm(n, mean = 0, sd = 1)
X = Ex
Z = 2 * X + Ey
Y = 0.5 * Z + Ez

library(dagitty)
d <- dagitty('dag {
             X -> Z
             Z -> Y
}')
plot(d)

dat <-  data.frame(Ex = Ex, Ey = Ey, Ez = Ez, X = X, Y = Y, Z = Z)

fit1 <- lm(Y ~ X, data = dat)
summary(fit1)
confint(fit1)
```
```{r}
fit2 <- lm(Y ~ X + Z, data = dat)
summary(fit2)
confint(fit2)
```
```{r}
mean(dat$Y)
```


```{r}
t <-  t.test(dat$Y,conf.level = 0.95)
t
```
# Exercise 3
```{r}
d <- dagitty('dag {
             X -> Z
             Y -> Z
             X -> Y
}')
plot(d)

```
```{r}
set.seed(124)
n <-  1000
Ex <- rnorm(n, mean = 0, sd = 1)
Ey <- rnorm(n, mean = 0, sd = 1)
Ez <- rnorm(n, mean = 0, sd = 1)

X = Ex

Y = 0.7 * X + Ey

Z = 0.8 * X + 0.5*Y + Ez



dat <-  data.frame(Ex = Ex, Ey = Ey, Ez = Ez, X = X, Y = Y, Z = Z)

fit3 <- lm(Y ~ X, data = dat)



confint(fit3)
```
