---
title:  "To adjust or not to adjust?"
author: "Beate Sick"
date:   "`r Sys.Date()`"
editor: source
lang: de
language: 
  title-block-author-single: "Autor:"
  title-block-published: "Version vom:"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    theme: cosmo
    fig-width: 8
    fig-height: 6
    fig-align: "center"
    fig-caption: true
    df-print: kable
    highlight: "monochrome"
---

```{css, echo = FALSE}
.justify {
  text-align: justify
}
```

In these tasks, the DAG and the SCM for the data-generating process involving three variables X, Y, Z are
given. The goal of the tasks is to check for different DAGs, if the total causal effect of X on Y can be
estimated by the unadjusted (y x) or the adjusted (y x + z) regression model.

## Exercise 1

**Pipe x -> z -> y**

Structural equation model:  
$X = \epsilon_X$  
$Z = 2 · X + \epsilon_Z$  
$Y = 0.5 · Z + \epsilon_Y$  

with $\epsilon_X ∼ N (0, 1)$, $\epsilon_Z ∼ N (0, 1)$ and $\epsilon_Y ∼ N (0, 1)$ being independent of each other.

### Teilaufgabe a
What is the total cause of X on Y

$Y = 0.5 · 2 = 1$ 

### Teilaufgabe b
Simulate n = 1000 data points. Then fit   
(1) a simple regression with X as the only predictor and   
(2) a multiple linear regression with both X and Z as predictors.   
In which of the fitted regression models is the true causal effect within the confidence interval of the effect of X on Y ? Set your seed to 123.

```{r, message=FALSE, warning=FALSE}
set.seed(123)
n <-  1000
Ex <- rnorm(n, mean = 0, sd = 1)
Ey <- rnorm(n, mean = 0, sd = 1)
Ez <- rnorm(n, mean = 0, sd = 1)
X = Ex
Z = 2 * X + Ez
Y = 0.5 * Z + Ey

library(dagitty)
d <- dagitty('dag {
             X -> Z
             Z -> Y
}')
plot(d)

dat <-  data.frame(Ex = Ex, Ey = Ey, Ez = Ez, X = X, Y = Y, Z = Z)

fit1 <- lm(Y ~ X, data = dat)
summary(fit1)
confint(fit1)
```
```{r, message=FALSE, warning=FALSE}
fit2 <- lm(Y ~ X + Z, data = dat)
summary(fit2)
confint(fit2)
```
```{r}
mean(dat$Y)
```


```{r}
t <-  t.test(dat$Y,conf.level = 0.95)
t
```

```{r, message=FALSE, warning=FALSE}
adjustmentSets(d, exposure = "X", outcome = "Y")

```
[The unadjusted regression model is sufficient to estimate the total causal effect of X on Y. In a pipe we should not adjust for the mediator. Note: there is no direct causal effect of X on Y.]{style="color:#00008B; font-weight:bold"}


## Exercise 2 Collider

Structural equation model:    
$X = \epsilon_X$  
$Y = 0.7 · X + \epsilon_Z$  
$Z = 0.8 · X + 0.5 · Y + \epsilon_Y$  

with $\epsilon_X ∼ N (0, 1)$, $\epsilon_Z ∼ N (0, 1)$ and $\epsilon_Y ∼ N (0, 1)$ being independent of each other.


```{r}
d <- dagitty('dag {
             X -> Z
             Y -> Z
             X -> Y
}')
plot(d)

```
### Teilaufgabe a

What is the total cause of X on Y?
0.7

### Teilaufgabe b

```{r}
set.seed(124)
n <-  1000
Ex <- rnorm(n, mean = 0, sd = 1)
Ey <- rnorm(n, mean = 0, sd = 1)
Ez <- rnorm(n, mean = 0, sd = 1)

X = Ex

Y = 0.7 * X + Ey

Z = 0.8 * X + 0.5*Y + Ez



dat <-  data.frame(Ex = Ex, Ey = Ey, Ez = Ez, X = X, Y = Y, Z = Z)

fit3 <- lm(Y ~ X, data = dat)



confint(fit3)
```

```{r}
adjustmentSets(d, exposure = "X", outcome = "Y")
```


[The unadjusted regression model is sufficient to estimate the total causal effect of X on Y. In a collider we should not adjust for the collider.]{style="color:#00008B; font-weight:bold"}